---
title: "Tidy Data"
author: "J. Zeldin"
date: "September 14, 2018"
output: html_document
---


## Teaching Modules

### I. Working with Tidy data

This module is an introduction to working with "tidy" data. This is a specific way of organizing data in R and leverages the tidyverse family of packages (Wickham et. al.) which include 'ggplot2', 'dplyr', 'tidyr', and many others. This is just an introduction, read Wickham's "R for Data Science" for more in-depth details of what tidy data is and the associated work-flow. 

**The three rules that make data tidy as are as follows:**  
1. Each variable must have its own column.  
2. Each observation must have its own row.  
3. Each value must have its own cell.  


Working with data in this format facilitates all of the tidyverse functions and manipulations you will need for exploration and analysis.  

Often, the data you would like to work with is not in a tidy format by default. There are a number of functions in the tidyr package that help to conform your data to the tidy format. Here's an example of an un-tidy dataset which we will make tidy with tidyr functions:



```{r, message=FALSE,warning=F}
library(tidyverse)
```

```{r, message=FALSE}
# This is our un-tidy dataset. We will import it using read_csv() (the tidyverse verison of read.csv)

teachers <- read_csv("teachers.csv")
head(teachers,15)
```

The dataset above is NOT tidy! We know this because it breaks our 2nd rule: Each observation must have its own row. In this case, each county is an observation and in our dataset there are two rows per county. How do we fix this to make our data tidy?

```{r}
# To make our dataset tidy, we will use the spread() function to spread the school column into two seperate variables with values from the count column.

# The spread function has two arguments:
# 1) the "key" argument, which takes the column with the variable names that are to be spread. In our case, this is the school column.
# 2) the "value", which takes the column containing the values associated with the variables that are to be spread. In our case, this is the count column.

teachers %>% 
  spread(key = school, value = count)

```
What if we had a dataset the looks like the one below? It is even more un-tidy! This time, in addition to haveing multiple rows for each observation, we have a situation where each variable does not have its own column. That is, county and state are combine in a single column. We can tidy this data with the separate() function

```{r}
teachers2 <-  read.csv("teachers2.csv")
head(teachers2,15)

teachers2 %>% 
  spread(key = school, value = count) %>% # take care of multiple rows per
                                          # observatuion, as above
  separate(state_county,into = c("state","county"),sep = "-")
                                          # here we used separate to split
                                          # state and county, using the 
                                          # separator "-" to tell the                                               # function where to split

  

```

There are many more functions for data wrangling in the tidyverse. Explore them to more efficiently handle and explore your data.



